---
title: "Simulating a t-Statistic"
output: html_notebook
---

In this activity, we'll simulate the distribution of the t-statistic.

First, let's assume that we know the true population standard deviation.  For this exercise, it will always be 5.  The following function takes a sample and computes a z-statistic for a reference mean of 0, using the true population standard deviation.

```{r}
statistic = function(sample){
  sample_mean = mean(sample)
  n = length(sample)
  return(sample_mean / (5 / sqrt(n)))
}
```

Next, the following code simulates an experiment many times.  In each experiment, a sample of size n is drawn from a normal distribution with mean 0 and standard deviation 5.  The a statistic is calculated using the above function.  Finally, the results are plotted against the standard normal curve, in red.

```{r}
library(ggplot2)
n = 7
data = data.frame(statistic = replicate(100000, statistic(rnorm(n, 0, 5))))
ggplot(data, aes(x = statistic)) + 
    geom_histogram(aes(y =..density..),
                   breaks = seq(-3, 3, by = .1), 
                   colour = "black", 
                   fill = "white") +
   xlim(c(-3,3)) +
stat_function(fun = dnorm, color = 'red')
```

1. Change the statistic function above so that it does not use the true population standard deviation (which a researcher never knows), but instead uses the sample standard deviation.

2. Compare the resulting distribution against the normal curve.  See how the difference changes if you increase n and decrease n.

