---
title: "hypothesis_test_practice_activity"
author: "Alex Hughes"
date: "10/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package loads, message=FALSE}
library(tidyverse)
library(magrittr)
```

In this short activity, you're going to write, and execute a short series of hypothesis tests using the `R` estimating framework. The data that you will be using comes from *Tidy Tuesday* and is about the costs of University education for undergraduates. 

This data collection effort is described in detail [on Github](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-10/readme.md). What we're going to ask you do start with is a very constrained set of tests that you might conduct with this data. If you're interested, there is much **much** more that you could investigate! 

```{r read data} 
base_url <- 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/'

tuition_cost       <- read_csv(file = paste0(base_url, 'tuition_cost.csv'))
tuition_income     <- read_csv(file = paste0(base_url, 'tuition_income.csv'))
salary_potential   <- read_csv(file = paste0(base_url, 'salary_potential.csv'))
historical_tuition <- read_csv(file = paste0(base_url, 'historical_tuition.csv'))
diversity_school   <- read_csv(file = paste0(base_url, 'diversity_school.csv'))
```
```{r}
head(tuition_income)
```

# One-Sample t-test 

What was the cost of tuition for the 1985-1986 school year, in current dollars, for public schools? 

```{r}
## this query will return a list of the costs for all institution types. 
historical_tuition %>% 
  filter(
    year == '1985-86' & 
      tuition_type == 'All Current')

## add a filter to this query to return only the rows for public schools 
## once you're able to add this filter, also add a `select()` verb to your 
## code (on a new line, after a new pipe) so that saved into the object 
## public_school_cost_1985 is only `tuition_cost`.  

public_school_cost_1985 <- historical_tuition %>% 
  filter(
    year == '1985-86' & 
      tuition_type == 'All Current' & 
      type == 'Public') %>% 
  select(tuition_cost) %>% 
  as.numeric() # This last line is to store the value outside a data.frame.

## so that you can check your results, the value should be $3571.
```

```{r}
tuition_cost %>% 
  filter(type == 'Public') %$% ## this 'pipe-like' operator comes from magrittr
                               ## and exposes the variable, outside a data.frame
  t.test(in_state_tuition, mu = public_school_cost_1985)

?%$%
```

# Two-Sample t-test 

# Paired t-test 

# Wilcox Rank Sum 

# Wilcox Sign Rank 

