---
title: "Visualizing Confidence Intervals"
output: html_document
---

```{r load packages}
library(dplyr)
library(ggplot2)
```


In this activity, we'll simulate a set of experiments, drawing data repeatedly from a distribution, then visualize the confidence intervals that result.

First, recall the equation for a 95% confidence interval:

$$
  [
    \overline{X} - \frac{s}{\sqrt{n}}t_{.025,n-1}, 
    \overline{X} + \frac{s}{\sqrt{n}}t_{.025,n-1}
  ]
$$

Read each of the lines in the following function, and note how they map from the math above, into applying the math in this function. 

```{r fill out confidence interval function}
confidence_interval = function(sample) {
  
  sample_mean    <- mean(sample)
  sample_sd      <- sqrt(var(sample))
  n              <- length(sample)
  critical_value <- qt(0.025, df=n-1)
  
  low_ci  <- sample_mean + (sample_sd / sqrt(n)) * critical_value
  high_ci <- sample_mean - (sample_sd / sqrt(n)) * qt(0.025, df=n-1)

  return(
    c(low_ci, high_ci)
    )      
}
```

```{r}
one_sample <- rnorm(n = 5, mean = 0, sd = 5)
confidence_interval(one_sample)
```


Next, run the following code to simulate an experiment many times.  

- In each experiment, a sample of size `n` is drawn from a normal distribution with mean 0 and standard deviation 5.
- Using the `confidence_interval` function you filled in, we produce 95% confidence intervals. 
- Finally, we plot these intervals. 

```{r make plots}

## There is no code for you to change in this plot. 

run_plots <- function(number_of_units = 5) { 
  ## This function takes an argument that is the number of draws for the population
  ## and returns a plot for the confidence intervals generated from 100 simulated
  ## samples of that size. 
  
  confidence_intervals <- data.frame(
    'low'      = rep(NA, 100), 
    'high'     = rep(NA, 100), 
    'captured' = rep(NA, 100)
    )
  
  for(i in 1:100) {
    our_sample <- rnorm(n = number_of_units, mean = 0, sd = 5)
    confidence_intervals[i, c('low', 'high')] <- confidence_interval(our_sample)
  }
  
  confidence_intervals <- confidence_intervals %>% 
    mutate(captured = low < 0 & high > 0)
  
  plot(
    NULL, 
    type="n", pch=19,
    xlim=c(1,100),
    ylim=c(min(confidence_intervals$low), max(confidence_intervals$high)),
    xlab="Trial", ylab=expression(mu))
  abline(
    h = c(0, qt(0.975, number_of_units-1)/sqrt(number_of_units), 
          qt(0.025, number_of_units-1)/sqrt(number_of_units)), 
    lty = c(1,2,2), 
    col = "gray")
  
  points(confidence_intervals$high, col = 2+confidence_intervals$captured, pch = 20)
  points(confidence_intervals$low,  col = 2+confidence_intervals$captured, pch = 20)

  for(i in 1:100) {
    lines(
      c(i,i), c(confidence_intervals$low[i],confidence_intervals$high[i]), 
      col = 2+confidence_intervals$captured[i], pch = 19)
  }
  
  title("Simulation of t-Confidence Intervals for the Population Mean")
  legend(
    0,-.75, 
    legend = c(
      expression(paste(mu," Captured")),
      expression(paste(mu," Not Captured"))), 
    fill = c(3,2)
    )
}
```

```{r}
run_plots()
```

Try to answer this question in your own words: what does 95% confidence mean?